---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Insert your name here"
date: "Insert date here"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

```{r start}
library(usethis)
use_git_config(
  user.name = "Rachel Weiner",
  user.email = "weinra22@wfu.edu"
)
```

```{r staff}
staff <- read_csv("data/instructional-staff.csv")
```
### Practice

```{r pivot-staff}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
staff_long
```

```{r plot-staff}
staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()
```
```{r line-plot-staff}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  labs(title = "Percentages of Faculty Hired by Type", color = "Faculty Type")
```

### Exercise 1

```{r load-Whichham}
data(Whickham)
```

This data set includes information regarding individuals health outcome as either "alive" or "dead" as well as their age and whether or not they smoke. As the data relies on current smoking behavior, this must be an observational study because it is unethical to assign experimental groups a "smoking treatment" if they were not already smokers. This study simply collected information on both individuals who smoke and who do not, without manipulating the groups themselves but simply observing already present behaviors. 

### Exercise 2

There are 1314 observations in this data set. Each observation represents an individual participant from the study. 

### Exercise 3

There are three variables in this data set, outcome, smoker, and age. The outcome variable regarding whether the individual is alive or dead and the smoker variable regarding whether the individual smokes (or smoked) are categorical variables, as well as binary variables as there is only two possible responses for each. For outcome, "dead" and "alive" are the two possible responses where as "yes" and "no" are the only possible responses for smoker. These variables are visualized below. 

```{r visualize-outcome}
Whickham %>%
  ggplot(aes(x = outcome)) +
  geom_bar()
```
Above is the visualized variable of outcome, as the individual is either alive or dead. The majority of respondents are alive. 

```{r}
Whickham %>%
  ggplot(aes(x = smoker)) +
  geom_bar()
```
Above is the visualized variabl of smoker as either someone who smokes (yes) or someone who does not (no). The majority of participants do smoke, but a similar but slightly less amount of participants do not smoke. 

```{r}
Whickham %>%
  ggplot(aes(x = age)) +
  geom_histogram()
```
Above is a histogram visualizing the age of participants and the distribution of their ages. As we can see, there is a relatively equal distribution of ages across all participants but there is a slightly greater amount around the age of 30. 

### Exercise 4

I would expect the relationship between smoking status and health to be negatively correlated. An individual that smokes is more likely to have a poor health outcome compared to individuals who do not smoke. 

### Exercise 5

```{r}
Whickham %>%
  count(smoker, outcome)
```

```{r}
contingency_table_2 <- table(Whickham$smoke, Whickham$outcome)

ftable(contingency_table_2)
```

```{r}
Whickham %>%
  ggplot(aes(x = outcome, fill = smoker)) + 
  geom_bar() +
  labs(title = "Health Outcomes by Smoking Status")
```
The bar plot above plots the health outcome of participants by smoking status. As we can see, about an even amount of alive participants smoke and slightly more of the dead particpants do not smoke. This visualization does not support my expected relationship between these variables as there is no clear effect of smoking on health outcome. 

### Exercise 6

Now I will create a new age variable that puts the participants into three age categories. The three categories are "18-44", "45-64", and "65+".

```{r}
Whickham <- Whickham %>%
  mutate(age_cat = case_when(age <= 44 ~ "18-44",
  age > 44 & age <= 64 ~ "45-64", age > 64 ~ "65+"))
```

### Exercise 7

```{r}
contingency_table_3 <- table(Whickham$age_cat, Whickham$smoke, Whickham$outcome)

ftable(contingency_table_3)
```
```{r}
ggplot(Whickham, aes(x = outcome, fill = smoker)) +
  geom_bar(position = "fill") +
  facet_wrap(~age_cat) +
  labs(title = "Health Outcome by Smoking Status",
       subtitle = "Faceted by Age Category")
```
Now that we have faceted the visualization by age category, we are able to see that there is a general 
